A heterogeneous landscape of selection and epistasis in protein-coding genes revealed by two-locus statistics

### 2022-03-30 manuscript/paper.tex
super close

Pairs of missense mutations that both fall within the same functional domain
have large positive LD that is elevated above that of pairs of synonymous
mutations that both fall within the same domain
(Figures~\ref{fig:data}C--F~and~\ref{fig:LDwithin}--\ref{fig:domainsEAS}). This
difference between pairs of missense and synonymous variants within domains is
especially pronounced for both linked pairs (\(\lesssim\) a few hundred base
pairs) (Figures~\ref{fig:data}E~and~\ref{fig:LDwithin}). Assuming most
synonymous mutations are neutral, we would expect signed LD between missense
mutations to be less than that between synonymous mutations under predictions
from models of either Hill-Robertson interference, synergistic epistasis, or
dominance (Figures~\ref{fig:epistasis}~and~\ref{fig:dominance}). This
observation is opposite to those expectations, suggesting a different
prevailing interactive effect between nonsynonymous mutations within domains
than those that are commonly tested or modeled.

The strength of selection on missense mutations within and outside of domains
is observed to differ, leading to an excess of rare missense mutations within
conserved domains (Table~\ref{tab:tajimasD}). LD is known to be sensitive to
allele frequencies, with rare mutations showing large positive signed LD
\citep{Good2022-ot}. To test whether the signal of increased LD between
missense mutations within domains is driven by rare variants, I considered
subsets of pairs of mutations based on their derived allele frequences
(Figures~\ref{fig:data}D~and~\ref{fig:domainsAFR}--\ref{fig:domainsEAS}). Rare
and uncommon variants show large average LD for each class of mutations, but
common variants recapitulate the opposing patterns of LD that is seen when
considering pairs at all frequencies. This is because unconditioned statistics
in the form of \(\sigma_d^1\) and \(\sigma_d^2\) are already dominated by
common variants.

@@

The large positive signal of LD for missense mutations within the same domain
does not extend to pairs of missense mutations that are in different domains.
Pairs of missense and synonymous mutations show nearly equal levels of LD close
to zero across domains, with missense mutations slightly more negative than
synonymous mutations (Figure \ref{fig:between_domains}). The interactive effect
driving large LD within domains is therefore likely domain-specific. However,
the average distance between pairs of mutations within domains is much smaller
than between domains, so this observation may be primarily driven by the higher
recombination distances between pairs of mutations across distinct domains.

Pairs of mutations that both fall outside of annotated domains have the
opposite pattern of signed LD to pairs of mutations falling within the same
domain. For pairs of mutations outside of domains but with distances matched to
those within domains, pairs of synonymous mutations have larger positive LD
than missense mutations. More distant pairs of mutations outside of domains,
matched to the same distances as the between-domain comparison, each have LD
roughly equal to zero.

The role that tightly linked variants have in driving these opposing signals
can be seen in the decay of signed LD with distance between mutations
(Figures~\ref{fig:LDgene}--\ref{fig:LDoutside}). Both synonymous and missense
mutation pairs at distances greater than a few hundred base pairs have average
LD that fluctuates around zero. However, for mutations outside domains,
synonymous variants separated by short distances have large positive LD, while
missense mutations have much lower LD (Figure~\ref{fig:LDoutside}). In contrast,
for mutations within the same domains, missense mutations have more positive
LD at short distances than synonymous mutations (Figure~\ref{fig:LDwithin}).

@@

By taking broad genome- and gene-wide surveys of LD across functional classes
of mutations, heterogeneous patterns of interactive effects that occur within
genes can be missed. From gene-wide averages, LD between pairs of missense
mutations does not appear to be different from pairs of synonymous variants,
although LD between loss-of-function variants is more negative. Hill-Robertson
effects are expected to be strongest for slightly to moderately deleterious
variants (with \(|s|\sim 1/N_e\)), as strongly deleterious mutations are not
expected to interfere with one another \citep{McVean2000-ox}. However,
loss-of-function mutations do not typically fall within this regime, and
inference of the distribution of fitness effects (DFE) for new loss-of-function
variants shows that a large majority are strongly deleterious (Supporting
Information). Instead, negative synergistic epistasis between strongly
deleterious mutations does produce large negative deviations of mean LD.
Strongly deleterious \emph{recessive} mutations can also produce this pattern,
making it difficult to distinguish these effects with certainty. This
difficulty is compounded by the large measurement noise for \(\E[D]\),
especially for loss-of-function variants for which only a few hundred
within-gene pairs exist in the human population data analyzed here.

@@

Positive average LD between both missense and synonymous mutations has
previously been reported in humans, drosophila, and other species
\citep{Sohail2017-zq,Garcia2021-zn,Sandler2021-of}. The similarity of their
gene-wide LD (Figure~\ref{fig:data}) might suggest that interference or
interactions between missense mutations are minimal, or at least no stronger
than those between synonymous mutations. However, averaging over all observed
pairs of mutations within a gene masks element-specific interactive effects
that drive LD in opposite directions. Nonsynonymous mutations found within
conserved protein domains, identified as conserved subunits of a gene with some
structural or functional role \citep{Stanek2020-pa}, are more strongly selected
against on average, but also have increased signed LD over synonymous mutations
at the same distances within domains. Missense mutation outside of domains but
at the same distances as those within domains have more negative LD, both
compared to distance-matched synonymous mutations outside of domains and to
mutations within domains. Neither dominance effects, synergistic epistasis, nor
Hill-Robertson interference are expected to result in positive LD, so some
other interactive effect should be driving this positive LD signal within
conserved domains.

@@

We do not see an increase in LD between mutations that are found in different
annotated domains within the same gene, which are on average at much larger
recombination distances. Nor do we find increased LD outside of domains.
Rather, for pairs of mutations outside of domains but matched to the same
distance as those within the same domain, missense variants have considerably
lower LD than synonymous variants. This difference between synonymous and
missense pairs of variants largely disappears for SNPs separated by more than a
few hundred base pairs (Figure~\ref{fig:data}E, F). This suggests that
Hill-Robertson interference is the primary mode of interaction between
mutations falling outside of domains, as epistasis is expected to impact LD
over larger distances than what is observed. Importantly, however, the strength
of epistasis is also likely to be a function of distance between mutations.

Taken together, selection on segregating variants within protein-coding genes
is nonuniform, with both the overall strength of selection and interactions
between variants differing between annotated elements. Missense and
loss-of-function mutations within conserved domains are subject to stronger
selection, skewing the SFS to lower frequencies. Typical approaches for
inferring the distribution of fitness effects from population genomic data
average over these differences by considering all nonsynonymous mutations
together \citep{Boyko2008-zk,Kim2017-xo}. If patterns of selective differ by
both mutational class and location or if loss-of-function variants are in fact
recessive, this can impact inferences of the DFE (Supporting Information). It
would be straightforward to adapt DFE-inference methods to infer more detailed
representation of the heterogeneous effects of new mutations within genes by
partitioning by missense and loss-of-function classes as well as by annotated
domains. Similarly, the different modes of interaction between mutations across
annotated domains mean that our standard models and simulation approaches may
be too simple to capture the evolutionary trajectories and patterns of
diversity that differ at finer scales.

@@

Spatial population structure may be responsible for the increase of LD observed
between synonymous variants, as every natural population lives in some
geography that will necessarily induce some structure. Neutral evolution in
structured populations alone cannot create this effect. Under standard
population genetic assumptions of constant mutation and recombination rates and
independent mutation events, migration, admixture, and spatial structure do not
result in nonzero \(\E[D]\). Rather, for a set of mutations with \emph{given}
differences in allele frequencies between source populations, \(D\) can be
nonzero in an admixed population between those particular mutations even when
\(D\) is zero in the source populations \citep{Cavalli-Sforza1971-jv}. This
result relies on conditioning on directional differences in allele frequencies
between populations. The expected difference in mean allele frequency between
source populations is zero (\(\E[(p_1-p_2)(q_1-q_2)] = 0\)), as differences are
equally likely to be positive or negative, and resulting \(\E[D]\) is still
zero when taking genome-wide averages.

@@

Nonrandom mutational processes provide an alternative explanation for positive
LD between neutral variants separated by short distances, in which clusters of
mutations occur simultaneously in the same mutational event. Such
multinucleotide mutations have been shown to affect patterns of LD over short
distances, on the order of 10s to 100s of base pairs \citep{Harris2014-zg}, and
clustered mutational events may be common in humans \citep{Besenbacher2016-ac}.
Indeed, a simple exponential model in which the fraction of pairs of mutations
arising from a multinucleotide mutation event decays with distance fits the
observed decay of \(\sigma_d^1\) between synonymous variants (Supporting
Information, Figure~\ref{fig:mslmnms}), requiring only a small fraction of
mutations to cause multinucleotide mutation events. We may therefore treat
positive LD as the baseline expectation for tightly linked variants primarily
due to clustered mutations, so that subsequent selective processes and
interactions cause LD to deviate from that expectation, as seen for missense
mutations across different sub-elements of genes. It may also be more
appropriate to compare the negative LD observed between linked loss-of-function
variants to that positive baseline expectation instead of zero, which would
imply that they are more recessive or have stronger synergistic epistasis than
from inferences assuming a neutral expectation of zero.

@@

When partitioning measurements of LD by mutation classes or regions within
genes, the decreasing number of pairwise comparisons leads to large estimated
measurement noise. Within each population, confidence intervals of observed
\(\sigma_d^1\) often overlap with zero or overlap with that of other classes of
mutations. While observed patterns are remarkably consistent across the 15
populations considered here, their joint evolutionary histories make formal
testing of significance difficult due to shared variation, so they cannot be
treated as independent measurements. Extensive forward simulations will likely
be required to more thoroughly assess significance.

@@

Instead, closed, numerical approaches to compute expectations for two-locus
statistics under a wide range of selection models, as presented here and in
\citet{Friedlander2022-bs}, allow us to explore this highly parameterized space
of models far more efficiently, and it opens the possibility for performing
likelihood-based inference using signed LD or other two-locus summaries of the
data. For example, inferring the joint distribution of dominance and selection
is underpowered using the SFS alone, but because signed LD is sensitive to the
levels of dominance (Figure~\ref{fig:dominance}), inferring the DFE with
dominance may be feasible using the joint distribution of allele frequencies
and LD.

@@

The two-locus sampling distribution is the direct analog to the single-locus
site frequency spectrum (SFS) of a given sample size. Instead of describing the
density or number of mutations with a given allele count out of \(n\) samples,
the two-locus distribution \(\Psi_n\) stores the density or number of pairs of
loci with observed haplotype counts, so that \(\Psi_n(i, j, k)\) is the number
of pairs for which we observe \(i\) copies of the \(AB\) haplotype, \(j\) of
type \(Ab\), \(k\) of type \(aB\), and \(n-i-j-k\) of type \(ab\). The size of
\(\Psi_n\) grows rapidly, with \(O(n^3)\) entries, which practically limits
computational approaches to moderate sample sizes \(n\lesssim 100\) and a
single population.

@@

Alternatively, \citet{Ragsdale2019-nt} showed that there exists a system of
ordinary differential equations directly on the entries of \(\Psi_n\). I
briefly summarize this general approach below, but refer readers to that
previous paper for detailed derivations of the drift, recombination, and
mutation terms and the moment-closure approximation. Instead, here I focus on
generalizing the selection operator to include epistasis, dominance, and other
forms of two-locus interactions.

@@

The moment system for \(\Psi_n\) can be derived directly from the diffusion
approximation, or it can be found through a more intuitive process of tracking
the dynamics of allelic states of a sample of size of \(n\) from the full
population. We assume \(n \ll N_e\), and \(r\) and \(s\) are \(O(1/N_e)\) so that
multiple coalescence, recombination, or selective events within the \(n\)
lineages are rare in any given generation. For example, for drift events, we
account for the probability that two lineages within the \(n\) tracked lineages
at time \(t+1\) coalesced in the previous generation and, given a coalescence
event, the probability that haplotype counts changed due to this coalescence
event {[}see Supporting Information; \citet{Jouganous2017-pq}; \citet{Ragsdale2019-nt}{]}. In
typical diffusion approximation fashion, we multiple through by \(2N_{ref}\) so
that time is measured in \(2N_{ref}\) generations, and we consider scaled
parameters \(\rho = 4Nr\), \(\theta = 4Nu\), and \(\gamma=2Ns\).

@@

In the absence of selection and for fully linked loci (i.e., \(\rho=0\)), the
system is closed and can be solved exactly. However, for nonzero recombination
or selection, the entries of \(\Psi_n\) rely on the slightly larger sampling
distributions with sample sizes \(n+1\) (for recombination and additive
selection) or \(n+2\) (for non-additive selection). This is because if a
recombination event occurs within one of \(n\) lineages being tracked by
\(\Psi_n\), we need to draw an additional lineage from the full population to
recombine with that chosen lineage, thus requiring \(\Psi_{n+1}^t\) to find
\(\Psi_n^{t+1}\). Selection events similarly require extra lineages from the full
population, which replace a chosen lineage that fails to reproduce with
probability proportional to its selection coefficient, assuming negative
selection.

@@

To include selection, we consider a model where we draw lineages uniformly from
the previous generation, but keep lineages with probability proportional to
their fitness. In the absence of dominance, selection reduces to a haploid
model, with acceptance and rejection probabilities depending on the fitnesses
of each haploid copy, where haplotype \(Ab\) has fitness \(1 + s_{A}\), \(aB\)
has fitness \(1 + s_{B}\), and \(AB\) has fitness \(1 + s_{AB}\). We assume the
doubly ancestral haplotype \(ab\) has fitness \(1\), so fitnesses are relative
to that of \(ab\) haplotypes. The standard multiplicative fitness function
assumes that \(s_{AB} \approx s_{A} + s_{B}\) (\(s^2\approx0\)), and a model
for epistasis can be written as \[s_{AB} = (s_{A} + s_{B}) (1 + \epsilon),\] so
that \(\epsilon > 0\) implies synergistic epistasis, while \(\epsilon < 0\)
implies antagonistic epistasis.

To obtain the recursion equation under selection we consider drawing \(n\)
lineages from generation \(t\), which has an expected sampling distribution of
haplotype counts given by \(\Psi_n^t\). However, assuming \(s\leq0\) for each
derived haplotype, each of those sampled lineages has probability of being
rejected equal to the selection coefficient assigned to its haplotype state. If
a lineage is rejected, a replacement is drawn from the full population. Under
the assumption that \(ns \ll 1\), the probability that more than one selection
event occurs in any given generation is negligibly small, so that the case of
multiple simultaneous rejections can be ignored. Then \(\Psi_n^{t+1}\) relies
only on \(\Psi_n^t\) and \(\Psi_{n+1}^t\) for additive selection. The full
selection operator \(\mathcal{S}\) for additive selection is given in the
Supporting Information.

To account for dominance within loci, or other general forms of two-locus
selection, the selection operator no longer reduces to individual haplotypes,
but instead we need to know the state of two-locus genotypes. For example, the
fitness of an individual carrying an \(Ab\) haplotype depends on whether their
second haplotype is \(ab\), \(Ab\), \(aB\), or \(AB\). We can therefore assign
a selection coefficient to each possible diploid configuration, \(s_{Ab/ab}\),
\(s_{Ab/Ab}\), and so on. Assuming that the doubly homozygous ancestral
\(ab/ab\) genotype has relative fitness 1, this gives nine possible unique
selection coefficients in the most general two-locus selection model. Note that
\(AB/ab\) and \(Ab/aB\) genotypes need not have the same selection coefficient,
which allows for simulation under a gene-based model of dominance (Table
\ref{tab:selmodels}).

The general selection operator follows the same approach as the haploid
selection operator with epistasis described above. Now, in the case of a
selection event rejecting a lineage within our tracked samples, we need to draw
not only the replacement lineage from the full population but also a second
haplotype from the full population to form the diploid genotype, as this
determines the probability that we reject the focal haplotype. We again assume
that \(ns \ll 1\) for all genotype selection coefficients, so that we may
assume at most a single selection event occurs in any given generation. This
means that to find \(\Psi_n^{t+1}\) under a general two-locus selection model,
which encompasses dominance within either locus, gene-based dominance, or a
combination or dominance and epistatic effects, we need \(\Psi_{n+2}^t\).
Again, a full derivation and expressions for the general selection operator are
given in the Supporting Information.

@@

I considered four variable population size histories, two simple toy models and
two inferred from human populations in African and Europe. For each size
history scenario, I tracked the evolution of \(\Psi_n(t)\) for varying
selection models, plotting the trajectories of \(\sigma_d^1\) and
\(\sigma_d^2\) over time (Figures~\ref{fig:toy}, \ref{fig:relate_dom}, and
\ref{fig:toy_sd2}--\ref{fig:relate_dom_sd2}). The selection strength at both
loci was fixed at either \(\gamma=-1, -2, and -10\), and recombination was set
to zero.

The simple size change models both had ancestral \(N_e=10,000\), with one a
\(3\)-fold population expansion that occurs 3,000 generations ago, and the
other a \(5\)-fold reduction 2,000 generations ago followed by a recovery to
its initial size \(1,000\) generations ago. The size histories for YRI and CEU
were inferred using \texttt{Relate} \citep{Speidel2019-nj} applied to the
phased autosomal data from \citet{1000_Genomes_Project_Consortium2015-zq},
using default parameters as recommended in the \texttt{Relate} tutorial.
\texttt{Relate} returns estimates of coalescence rates within specified time
bins, and population sizes are estimated as their inverses. Estimates using
\texttt{Relate} for population sizes in the very recent past (\(<3,000\) years,
or \(\approx 100\) generations) diverged, so I truncated the history over this
time period and assumed a constant size from the most recent non-diverged bin.

>>>>

Pairs of missense mutations that both fall within the same functional domain
have large positive LD that is elevated above that of pairs of synonymous
mutations that both fall within the same domain
(Figures~\ref{fig:data}C--F~and~\ref{fig:LDwithin}--\ref{fig:domainsEAS}). This
difference between pairs of missense and synonymous variants within domains is
especially pronounced for linked pairs within a few hundred base pairs of each
other (Figures~\ref{fig:data}E~and~\ref{fig:LDwithin}). Assuming most
synonymous mutations are neutral, we would expect signed LD between missense
mutations to be less than that between synonymous mutations under predictions
from models of either Hill-Robertson interference, synergistic epistasis, and
some models of dominance (Figures~\ref{fig:epistasis}~and~\ref{fig:dominance}).
This observation is opposite to those expectations, suggesting a different
prevailing interactive effect between nonsynonymous mutations within domains.

The strength of selection on missense mutations within and outside of domains
is observed to differ, leading to an excess of rare missense mutations within
conserved domains (Table~\ref{tab:tajimasD}). LD is known to be sensitive to
allele frequencies, with rare mutations showing large positive signed LD
\citep{Good2022-ot}. To test whether the signal of increased LD between
missense mutations within domains is driven by rare variants, I considered
subsets of pairs of mutations based on their derived allele frequences
(Figures~\ref{fig:data}D~and~\ref{fig:domainsAFR}--\ref{fig:domainsEAS}). Rare
and uncommon variants show large average LD for each class of mutations, but
common variants recapitulate the opposing patterns of LD that is seen when
averaging over pairs at all frequencies, as unconditioned statistics in the
form of \(\sigma_d^1\) and \(\sigma_d^2\) are dominated by common variants.

@@

The large positive signal of LD for missense mutations within the same domain
does not extend to pairs of missense mutations that are in different domains.
Pairs of missense and synonymous mutations show nearly equal levels of LD close
to zero across domains, with missense mutations slightly more negative than
synonymous mutations (Figure \ref{fig:between_domains}). The interactive effect
driving large LD in domains is therefore likely domain-specific. However, the
average distance between pairs of mutations within domains is much smaller than
between domains, and this observation may be primarily driven by the higher
recombination distances between pairs of mutations across distinct domains.

Pairs of mutations that both fall outside of annotated domains have the
opposite pattern of signed LD to pairs of mutations falling within the same
domain. For pairs of mutations outside of domains but with distances matched to
those within domains, pairs of synonymous mutations have larger positive LD
than missense mutations. More distant pairs of mutations outside of domains,
matched to the same distances as the between-domain comparison, each have LD
roughly equal to zero (Figure~\ref{fig:between_domains}).

The role that tightly linked variants have in driving these opposing signals
can be seen in the decay of signed LD with distance between mutations
(Figures~\ref{fig:LDgene}--\ref{fig:LDoutside}). Both synonymous and missense
mutation pairs at distances greater than a few hundred base pairs have average
LD that fluctuates around zero. However, for mutations outside domains,
synonymous variants separated by short distances have large positive LD, while
missense mutations have lower LD (Figure~\ref{fig:LDoutside}). In contrast, for
mutations within the same domains, missense mutations have more positive LD at
short distances than synonymous mutations (Figure~\ref{fig:LDwithin}).

@@

By taking broad genome- and gene-wide surveys of LD across functional classes
of mutations, heterogeneous patterns of interactive effects that occur within
genes can be missed. From gene-wide averages, LD between pairs of missense
mutations does not appear to be different from pairs of synonymous variants,
although LD between loss-of-function variants is more negative. Hill-Robertson
effects are expected to be strongest for slightly to moderately deleterious
variants (with \(|s|\sim 1/N_e\)), as strongly deleterious mutations are not
expected to interfere with one another \citep{McVean2000-ox}. However,
loss-of-function mutations do not typically fall within this regime, and
inference of the distribution of fitness effects (DFE) for new loss-of-function
variants shows that a large majority are strongly deleterious (Supporting
Information).

Instead, negative synergistic epistasis between strongly deleterious mutations
does produce large negative deviations of mean LD. Weakly deleterious
recessive mutations can also produce this pattern, but strongly deleterious
recessive mutations lead to slightly positive LD \citep{Roze2021-cf}. While
most loss-of-function mutations are strongly deleterious, those that rise to
appreciable frequency are likely more benign and \(\sigma_d^1\) may be driven
by patterns of weakly deleterious loss-of-function mutations. The difficulty in
distinguishing these effects is compounded by the large measurement noise for
\(\E[D]\), especially for loss-of-function variants for which only a few
hundred within-gene pairs exist in the human population data analyzed here and
which are separated by larger distances on average than neighboring missense
and synonymous mutations.

In addition to LD varying by distance, LD can also vary due to differences in
allele frequencies among classes of mutations (as selection drives mutations to
higher or lower frequencies). Matching distances between pairs as well as
allele frequecies between classes of mutations helps to reduce these concerns.
It is possible that recombination rates can vary between annotated regions,
resulting in differing patterns of background selection, which can affect
allele frequencies and LD (e.g., Figures~\ref{fig:bgs1}--\ref{fig:bgs4}). I did
not condition on local recombination rates or inferred levels of background
selection here, which is left to future work.

@@

Positive average LD between both missense and synonymous mutations has been
reported in humans, \emph{Drosophila}, and other species
\citep{Sohail2017-zq,Sandler2021-of}, while others have found that
nonsynonymous mutations show lower LD than synonymous mutations
\citep{Garcia2021-zn}. The similarity of their gene-wide LD observed in this
study (Figure~\ref{fig:data}) might suggest that interference or interactions
between missense mutations are minimal, or at least no stronger than those
between synonymous mutations. However, averaging over all observed pairs of
mutations within a gene masks element-specific interactive effects that drive
LD in opposite directions. Nonsynonymous mutations found within conserved
protein domains, identified as conserved subunits of a gene with some
structural or functional role \citep{Stanek2020-pa}, are more strongly selected
against on average, but also have increased signed LD over synonymous mutations
at the same distances within domains. Missense mutation outside of domains but
at the same distances as those within domains have more negative LD, both
compared to distance-matched synonymous mutations outside of domains and to
mutations within domains.  Neither dominance effects (aside from very strongly
selected recessive mutations \citep{Roze2021-cf}), synergistic epistasis, nor
Hill-Robertson interference are expected to result in positive LD, so some
other interactive effect should be driving this signal of positive LD within
conserved domains.

@@

We do not see an increase in LD between mutations that are found in different
annotated domains within the same gene, which are on average at much larger
recombination distances. Nor do we find increased LD outside of domains.
Rather, for pairs of mutations outside of domains but matched to the same
distance as those within the same domain, missense variants have considerably
lower LD than synonymous variants. This difference between synonymous and
missense pairs of variants largely disappears for SNPs separated by more than a
few hundred base pairs (Figure~\ref{fig:data}E, F). This suggests that
Hill-Robertson interference is the primary mode of interaction between missense
mutations falling outside of domains, in agreement with \citet{Garcia2021-zn},
as epistasis is expected to impact LD over larger distances than what is
observed. Importantly, however, the strength of epistasis is also likely to be
a function of distance between mutations, complicating this interpretation.

Taken together, selection on segregating variants within protein-coding genes
is nonuniform, with both the overall strength of selection and interactions
between variants differing between annotated elements. Missense and
loss-of-function mutations within conserved domains are subject to stronger
selection, skewing the SFS to lower frequencies. Typical approaches for
inferring the distribution of fitness effects from population genomic data
average over these differences by considering all nonsynonymous mutations
together \citep{Boyko2008-zk,Kim2017-xo}. If patterns of selection differ by
both mutational class and location or if loss-of-function variants are in fact
recessive, this can impact inferences of the DFE (Supporting Information). It
would be straightforward to adapt DFE-inference methods to infer more detailed
representation of the heterogeneous effects of new mutations within genes by
partitioning by missense and loss-of-function classes as well as by annotated
domains. Similarly, the different modes of interaction between mutations across
annotated domains mean that our standard models and simulation approaches may
be too simple to capture the evolutionary trajectories and patterns of
diversity that differ at finer scales.

@@

Spatial population structure may be responsible for the increase of LD observed
between synonymous variants. Neutral evolution in structured populations alone
cannot create this effect. Under standard population genetic assumptions of
constant mutation and recombination rates and independent mutation events,
migration, admixture, and spatial structure do not result in nonzero \(\E[D]\).
Rather, for a set of mutations with \emph{given} differences in allele
frequencies between source populations, \(D\) can be nonzero in an admixed
population between those particular mutations even when \(D\) is zero in the
source populations \citep{Cavalli-Sforza1971-jv}. This result relies on
conditioning on directional differences in allele frequencies between
populations. Without conditioning, the expected difference in mean allele
frequency between source populations is zero (\(\E[(p_1-p_2)(q_1-q_2)] = 0\)),
as differences are equally likely to be positive or negative, and thus
resulting \(\E[D]\) is still zero when taking genome-wide averages.

@@

Nonrandom mutational processes provide an alternative explanation for positive
LD between neutral variants separated by short distances, in which clusters of
mutations occur simultaneously in the same mutational event. Such
multinucleotide mutations have been shown to affect patterns of LD over short
distances, on the order of 10s to 100s of base pairs \citep{Harris2014-zg}, and
clustered mutational events may be common in humans \citep{Besenbacher2016-ac}.
Indeed, a simple exponential model in which the fraction of pairs of mutations
arising from a multinucleotide mutation event decays with distance fits the
observed decay of \(\sigma_d^1\) between synonymous variants (Supporting
Information, Figure~\ref{fig:mslmnms}), with the best-fit model requiring only
a small fraction of mutations to cause multinucleotide mutation events. We may
therefore treat positive LD as the baseline expectation for tightly linked
variants primarily due to clustered mutations, so that subsequent selective
processes and interactions cause LD to deviate from that expectation, as seen
for missense mutations across different sub-elements of genes. It may also be
more appropriate to compare the negative LD observed between linked
loss-of-function variants to that positive baseline expectation instead of
zero, which would imply that they are more recessive or have stronger
synergistic epistasis than from inferences assuming a neutral expectation of
zero. Additional analyses and extensive simulations will be required to tease
apart the effects of population structure, selective interference at linked
sites, and clustered mutational events on patterns of LD.

@@

When partitioning measurements of LD by mutation classes or regions within
genes, the decreasing number of pairwise comparisons leads to large estimated
measurement noise. Within each population, confidence intervals of observed
\(\sigma_d^1\) often overlap with zero or overlap with that of other classes of
mutations. While observed patterns are remarkably consistent across the 15
populations considered here, their joint evolutionary histories make formal
testing of significance difficult due to shared variation, as they cannot be
treated as independent measurements. Extensive forward simulations will likely
be needed to more thoroughly assess significance.

@@

Instead, closed, numerical approaches to compute expectations for two-locus
statistics under a wide range of selection models, as presented here and in
\citet{Friedlander2022-bs}, allow us to explore this highly parameterized space
of models far more efficiently, and it opens the possibility for performing
likelihood-based inference using signed LD or other two-locus summaries of the
data. For example, inferring the joint distribution of dominance and selection
is underpowered using the SFS alone, but because signed LD is sensitive to the
levels of dominance (Figure~\ref{fig:dominance}), inferring the DFE with
dominance may be feasible using the joint distribution of allele frequencies
and LD. The results presented in this paper likely do not cover the space of
all possible two-locus models, and other unexplored models may result in
similar patterns of signed LD. Comparisons to empirical data should therefore
be treated with caution, and additional careful modeling will allow a more firm
determination of the interactive effects causing observed patterns of
variation.

@@

The two-locus sampling distribution is the direct analog to the single-locus
site frequency spectrum (SFS) of a given sample size
(Figure~\ref{fig:TwoLocusFS}). Instead of describing the density or number of
mutations with a given allele count out of \(n\) samples, the two-locus
distribution \(\Psi_n\) stores the density or number of pairs of loci with
observed haplotype counts, so that \(\Psi_n(i, j, k)\) is the number of pairs
for which we observe \(i\) copies of the \(AB\) haplotype, \(j\) of type
\(Ab\), \(k\) of type \(aB\), and \(n-i-j-k\) of type \(ab\). The size of
\(\Psi_n\) grows rapidly, with \(O(n^3)\) entries, which practically limits
computational approaches to moderate sample sizes \(n\lesssim 100\) and a
single population.

@@

Alternatively, \citet{Ragsdale2019-nt} showed that there exists a system of
ordinary differential equations directly on the entries of \(\Psi_n\). I
briefly summarize this general approach below, but refer readers to that paper
for detailed derivations of the drift, recombination, and mutation terms and
the moment-closure approximation. Instead, here I focus on generalizing the
selection operator to include epistasis, dominance, and other forms of
two-locus interactions.

@@

The moment system for \(\Psi_n\) can be derived directly from the diffusion
approximation, or it can be found through a more intuitive process of tracking
the dynamics of allelic states of a sample of size of \(n\) from the full
population. We assume \(n \ll N_e\), and \(r\) and \(s\) are \(O(1/N_e)\) so
that multiple coalescence, recombination, or selective events within the \(n\)
lineages are rare in any given generation. For example, for drift events, we
account for the probability that two lineages within the \(n\) tracked lineages
at time \(t+1\) coalesced in the previous generation and, given a coalescence
event, the probability that haplotype counts changed due to this coalescence
event \citep[Supporting Information;][]{Jouganous2017-pq,Ragsdale2019-nt}. In
typical diffusion approximation fashion, we multiple through by \(2N_{ref}\) so
that time is measured in \(2N_{ref}\) generations, and we consider scaled
parameters \(\rho = 4Nr\), \(\theta = 4Nu\), and \(\gamma=2Ns\).

@@

In the absence of selection and for fully linked loci (i.e., \(\rho=0\)), the
system is closed and can be solved exactly. However, for nonzero recombination
or selection, the entries of \(\Psi_n\) rely on the slightly larger sampling
distributions with sample sizes \(n+1\) (for recombination and additive
selection) or \(n+2\) (for non-additive selection). This is because if a
recombination event occurs within one of \(n\) lineages being tracked by
\(\Psi_n\), we need to draw an additional lineage from the full population to
recombine with that chosen lineage, thus requiring \(\Psi_{n+1}^t\) to find
\(\Psi_n^{t+1}\). Selection events similarly require extra lineages from the full
population, which replace a chosen lineage that fails to reproduce with
probability proportional to its relative fitness.

@@

To include selection, we consider a model where we draw lineages uniformly from
the previous generation, but keep lineages with probability proportional to
their fitness. In the absence of dominance, selection reduces to a haploid
model, with acceptance and rejection probabilities depending on the fitnesses
of each haploid copy, where haplotype \(Ab\) has fitness \(1 + s_{A}\), \(aB\)
has fitness \(1 + s_{B}\), and \(AB\) has fitness \(1 + s_{AB}\). We assume the
doubly ancestral haplotype \(ab\) has fitness \(1\), so fitnesses are relative
to that of \(ab\) haplotypes. The standard multiplicative fitness function
assumes that \(s_{AB} \approx s_{A} + s_{B}\) (assuming \(s^2\approx0\)), and a
model for epistasis can be written as \[s_{AB} = (s_{A} + s_{B}) (1 +
\epsilon),\] so that \(\epsilon > 0\) implies synergistic epistasis, while
\(\epsilon < 0\) implies antagonistic epistasis.

To obtain the recursion equation under selection we consider drawing \(n\)
lineages from generation \(t\), which has an expected sampling distribution of
haplotype counts given by \(\Psi_n^t\). However, assuming \(s\leq0\) for each
derived haplotype, each of those sampled lineages has probability of being
rejected equal to the absolute value of the selection coefficient assigned to
its haplotype state. If a lineage is rejected, a replacement is drawn from the
full population. Under the assumption that \(ns \ll 1\), the probability that
more than one selection event occurs in any given generation is negligibly
small, so that the case of multiple simultaneous rejections can be ignored.
Then \(\Psi_n^{t+1}\) relies only on \(\Psi_n^t\) and \(\Psi_{n+1}^t\) for
additive selection. The full selection operator \(\mathcal{S}\) for additive
selection is given in the Supporting Information.

To account for dominance, or other general forms of two-locus selection, the
selection operator no longer reduces to individual haplotypes, but instead we
need to know the state of two-locus genotypes. For example, the fitness of an
individual carrying an \(Ab\) haplotype depends on whether their second
haplotype is \(ab\), \(Ab\), \(aB\), or \(AB\). We can therefore assign a
selection coefficient to each possible diploid configuration, \(s_{Ab/ab}\),
\(s_{Ab/Ab}\), and so on. Assuming that the doubly homozygous ancestral
\(ab/ab\) genotype has relative fitness 1, this gives nine possible unique
selection coefficients in the most general two-locus selection model. Note that
\(AB/ab\) and \(Ab/aB\) genotypes need not have the same selection coefficient,
which allows for simulation under a gene-based model of dominance (Table
\ref{tab:selmodels}).

The general selection operator follows the same approach as the haploid
selection operator with epistasis described above. Now, in the case of a
selection event rejecting a lineage within our tracked samples, we need to draw
not only the replacement lineage from the full population but also a second
haplotype from the full population to form the diploid genotype, as this
determines the probability that we reject the focal haplotype. We again assume
that \(ns \ll 1\) for all genotype selection coefficients, so that we may
assume at most a single selection event occurs in any given generation. This
means that to find \(\Psi_n^{t+1}\) under a general two-locus selection model,
which encompasses dominance within either locus, gene-based dominance, or a
combination of dominance and epistatic effects, we need \(\Psi_{n+2}^t\).
Again, a full derivation and expressions for the general selection operator are
given in the Supporting Information.

@@

I considered four variable population size histories, two simple toy models and
two inferred from human populations in African and Europe using \texttt{Relate}
\citep{Speidel2019-nj}. For each size history scenario, I tracked the evolution
of \(\Psi_n(t)\) for varying selection models, plotting the trajectories of
\(\sigma_d^1\) and \(\sigma_d^2\) over time (Figures~\ref{fig:toy},
\ref{fig:relate_dom}, and \ref{fig:toy_sd2}--\ref{fig:relate_dom_sd2}). The
selection strength at both loci was fixed at either \(\gamma=-1\) or \(-10\)
for the models with epistasis, or \(\gamma=-2\) for models with dominance, and
recombination was set to zero.
